---
title: "Predicting Health Insurance Costs"
author: "Zara Nip and Nehal Linganur"
output:
  pdf_document: default
---

### Dependencies
```{r}
library(NbClust)
library(mclust)
library(flexclust)
library(cluster)
library(fpc)
```

### Data Preparation
Import data and remove missing rows.
```{r}
insurance <- read.csv("insurance.csv")
insurance <- na.omit(insurance)
insurance$age <- as.numeric(insurance$age)
insurance$bmi <- as.numeric(insurance$bmi)

```

### Linear Regression Models

```{r}
mod<-lm(charges~., data=insurance)
summary(mod)
```
Finding residuals and QQ plots for all factors
```{r}
par(mfrow=c(1,2))
hist(mod$residuals, prob = TRUE)
lines(density(mod$residuals))

qqnorm(y=mod$residuals)
qqline(y=mod$residuals, datax = FALSE)
```


Using only statistically significant factors: age, bmi, children, smoker (yes).
```{r}
sigFac<-lm(charges~age + bmi + children + smoker, data=insurance)
summary(sigFac)
```
### Residual Plots
```{r}
par(mfrow=c(1,2))
hist(sigFac$residuals, prob = TRUE)
lines(density(sigFac$residuals))

qqnorm(y=sigFac$residuals)
qqline(y=sigFac$residuals, datax = FALSE)
```
### Testing Correlation With Clustering
Our previous model stated that age, bmi, children, smoker (yes) were all significant factors. We are now going to perform a clustering model to see if this still holds true.

```{r}
seed.val<-1234

insurance.scaled <- scale(insurance[1] + insurance[3]) 
for(i in 1:ncol(insurance.scaled)){
print(max(insurance.scaled[ , i]))
}

```
```{r}
RNGversion("4.1.2")
set.seed(seed.val)


clusterInsurance <- kmeans(insurance.scaled, 2, nstart = 25)
```

```{r}
clusterInsurance$size
clusterInsurance$tot.withinss
clusterInsurance$betweenss
```

```{r}
wssplot <- function(data, nc=15, seed=1234){
               wss <- (nrow(data)-1)*sum(apply(data,2,var))
               for (i in 2:nc){
                    set.seed(seed)
                    wss[i] <- sum(kmeans(data, centers=i)$withinss)}
                plot(1:nc, wss, main = "Finding Optimal Number of Clusters using Within Group Sum of Squares", type="b", xlab="Number of Clusters",
                     ylab="Within groups sum of squares")
}

wssplot(insurance.scaled)
```

```{r}
plotcluster(insurance.scaled, clusterInsurance$cluster)
title("Graphical Representation of Clusters")
```
Commented out because of knitting issues. Returned k = 2 as optimal amount of clusters.

#nc <- NbClust(insurance.scaled, min.nc=2, max.nc=15, method="kmeans")

#table(nc$Best.n[1,])

Optimal number of clusters using PAM
```{r}
dist.mat<-daisy(insurance.scaled,  metric="euclidean")
pk <- pamk(dist.mat, krange=2:15, usepam=TRUE, diss=TRUE)
pk$nc
```

```{r}
fit.pam = pam(dist.mat,2)
plot(fit.pam)

jpeg("MYPLOT.jpg")
plot(fit.pam)
dev.off()
```
Chi-Squared Test
```{r}
cont.table <- table(insurance$charges, clusterInsurance$cluster)
print(chisq.test(cont.table))
```
How different are the clusters?
```{r}
randIndex(cont.table)

```

